Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-17 ind_files/01_Introduction.tex
Minor phrasing changes

Studies of neutral or near-neutral dynamics in population and quantitative genetics usually do take stochasticity seriously, explicitly modeling finite populations that follow stochastic dynamics. Unfortunately, the classic or standard stochastic models in both population genetics~\citep{fisher_genetical_1930,wright_evolution_1931, moran_random_1958, kimura_diffusion_1964} and quantitative genetics~\citep{crow_introduction_1970, lande_natural_1976} typically assume a fixed total population size and their validity is therefore rather restrictive since population sizes routinely fluctuate in the real world. Those models which study evolution in populations with non-constant size usually impose deterministic, usually phenomenological, rules that are independent of population composition for how the total population size must vary~\citep{kimura_probability_1974,ewens_probability_1967,otto_probability_1997, engen_fixation_2009, waxman_unified_2011}. Such models are somewhat artificial since demography and population size are forced to be independent quantities even though this is obviously not the case in natural populations or individual-based/agent-based simulations, where population size is a `bulk' property whose value emerges from an intricate interplay of the individual-level demographic processes of birth and death. Notably, the Price equation itself is usually formulated in a deterministic, dynamically-insufficient manner (but see~\cite{rice_stochastic_2008} for a stochastic formulation of the Price equation). Since real-life populations are stochastic, finite, and of non-constant population size, this is somewhat of a problem, since we know that such deterministic approximations that may not capture important dynamics of the real system of interest - finite, fluctuating populations that vary stochastically.

@@

\section{A very brief outline of the rest of this thesis}
This section provides a bullet-point chapter-wise outline for those who are really short of time. A more detailed outline of the thesis is provided in the next section.

@@

\begin{itemize}
		\item Chapter \ref{chap_math_background} provides the necessary mathematical background and provides a toy example studying the size of a population of identical individuals that illustrates the major ideas used. If I am required to shoehorn the structure into the `intro-methods-results-discussion' format, then this chapter can be thought of as providing a mathematical `introduction' and illustration of the `methods' that will be used/generalized to get `results' in the following chapters.
		\item Chapter \ref{chap_BD} develops a formalism for describing the evolution of finite fluctuating populations of individuals that come in arbitrarily many `types' which vary in a \emph{discrete} character. This yields equations that generalize the classic Price equation and replicator-mutator equation to finite fluctuating populations.
		\item Chapter \ref{chap_infD_processes} extends the ideas developed in chapter \ref{chap_BD} to populations that vary in a single one-dimensional \emph{quantitative} character and derives a so-called `stochastic field theory' that describes evolution in such populations. This also results in some mathematical equations that may be of independent interest to physicists and applied mathematicians.
		\item Chapter \ref{chap_unification} provides a technical summary of the major results by deriving three important stochastic differential equations. These are equations for type frequencies, the population mean value of an arbitrary type-level quantity, and the population variance of an arbitrary type-level quantity, and respectively generalize the replicator-mutator equation, the Price equation, and \cite{lion_theoretical_2018}'s variance equation to finite, stochastically fluctuating populations. These equations predict a directional evolutionary force called `noise-induced selection' that is only seen in finite, fluctuating populations. Some implications for social evolution and community ecology are discussed. I also briefly discuss the field equation formalism I develop in chapter \ref{chap_infD_processes}.
		\item Chapter \ref{sec_disc} provides a quick summary of the major results and discusses biological implications, connections with previous studies, and opportunities for future work. This chapter has no equations (!). Thus, readers who do not want any mathematical equations can skip all other chapters and directly read chapter \ref{sec_disc} if they are interested only in the final results and takeaways. 
\end{itemize}

\section{A more expanded outline of the rest of this thesis}

This section provides a more detailed outline of the thesis including the technical aspects covered, which sections are original, etc.

@@

While the mathematics of chapter \ref{chap_BD} is standard and well-understood, it has, to the best of my knowledge, not been used in this context in the generality we use here. Several specific models of specific systems do use these mathematical techniques, but these papers are often written assuming familiarity with notions in physics and/or mathematics and thus may not be very accessible to theoretical ecologists who do not have formal training in these subjects (but see~\cite{czuppon_understanding_2021} for a recent pedagogical review on the general approach as applied to Wright-Fisher and Moran processes, where total population size is constant). As such, chapters \ref{chap_math_background} and \ref{chap_BD} together also serve as a tutorial and technical introduction to some theoretical ideas: For ecologists, the chapter introduces `system size expansions' and illustrates their use in a general setting, and can be seen as a tutorial on modelling finite populations analytically with minimal assumptions (Also see~\cite{czuppon_understanding_2021} for a similar review that works with tools mostly used in the more `mathematical' literature; I work here with tools mostly used in the physics literature\footnote{Though I do not typically use physics language/notation in this thesis because I am personally closer to mathematics than physics.}); For population geneticists, the chapter illustrates how system-size approximations (`diffusion approximations' in the population genetics literature) can be carried out without assuming a constant (effective) population size and how this generalization has important consequences for the evolutionary forces at play; For physicists and applied mathematicians, the chapter presents a study of the consequences of applying the system-size expansion to the kind of density-dependent birth-death processes that are widely applicable in ecology and evolution - Unlike many physical systems, though demographic processes like birth and death over ecological timescales are usually formulated in terms of population numbers or densities, predictions in evolution are typically in terms of frequencies of types, and this fact has subtle consequences that can be overlooked if one only works with densities.

Chapter \ref{chap_infD_processes} introduces a function-valued process\footnote{This is really something called a measure-valued process, but in the interest of making this thesis accessible to a broader audience, I will not use measure-theoretic language or tools henceforth.} to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via an analog of the system-size approximation to arrive at a `functional' Fokker-Planck equation in which derivatives are replaced by functional derivatives. I show that classic equations from quantitative genetics such as Kimura's infinite alleles model and Lande's gradient dynamics can be derived as the infinite population limit of this stochastic process. I also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation that can be analyzed for specific systems as required. Unlike the systems studied in Chapter \ref{chap_BD}, formalizing the study of the kind of processes we study in Chapter \ref{chap_infD_processes} is an active area of mathematical research~\citep{carmona_stochastic_1999,da_prato_stochastic_2014,prevot_concise_2007,liu_stochastic_2015,bogachev_fokker-planck-kolmogorov_2015,balan_gentle_2018} and the mathematics itself is far from settled.

>>>>

Studies of neutral or near-neutral dynamics in population and quantitative genetics usually do take stochasticity seriously, explicitly modeling finite populations that follow stochastic dynamics. Unfortunately, the classic or standard stochastic models in both population genetics~\citep{fisher_genetical_1930,wright_evolution_1931, moran_random_1958, kimura_diffusion_1964} and quantitative genetics~\citep{crow_introduction_1970, lande_natural_1976} typically assume a fixed total population size and their validity is therefore rather restrictive since population sizes routinely fluctuate in the real world. Those models which study evolution in populations with non-constant size usually impose deterministic, usually phenomenological, rules that are independent of population composition for how the total population size must vary~\citep{kimura_probability_1974,ewens_probability_1967,otto_probability_1997, engen_fixation_2009, waxman_unified_2011}. Such models are somewhat artificial since demography and population size are forced to be independent quantities even though this is obviously not the case in natural populations, where population size is a `bulk' property whose value emerges from an intricate interplay of the individual-level demographic processes of birth and death. Notably, the Price equation is usually formulated in a deterministic, dynamically insufficient\footnote{For the non-theorists: `Dynamically insufficient' means that the equation cannot be iterated in the form of a difference equation $x_{t+1} = F(x_t)$ or a differential equation $\dot{x} = F(x)$  to predict a dynamic `trajectory' given an initial condition $x_0$. The term `insufficient' is to indicate that the equation requires `complete' information and only relates two quantities `in retrospect': In our case, in its most general setting, the Price equation is usually formulated in a way that partitions a \emph{given} amount of phenotypic change between two populations (usually the same population at two different times) into change due to selection, transmission bias, etc., rather than \emph{predicting} a trajectory for how much phenotypic change will occur at various future times based on the \emph{current} phenotypic distribution.} setting (but see~\cite{page_unifying_2002} and~\cite{lion_theoretical_2018} for a dynamically sufficient formulation that is deterministic and~\cite{rice_stochastic_2008} for a stochastic formulation that is dynamically insufficient). Since real-life populations are stochastic, finite, and of non-constant population size, this is somewhat of a problem, since we know that such deterministic approximations that may not capture important dynamics of the real system of interest - finite, fluctuating populations that vary stochastically.

@@

\section{A very brief outline of the rest of this thesis}
This section provides a bullet-point chapter-wise outline for convenience. A more detailed outline of the thesis, with explanations of the technical content covered, originality, etc., is provided in the next section.

@@

\begin{itemize}
		\item Chapter \ref{chap_math_background} provides the necessary mathematical background and provides a toy example studying the size of a population of identical individuals that illustrates the major ideas used. If I am required to shoehorn this thesis into a `intro-methods-results-discussion' format, then chapter \ref{chap_math_background} can be thought of as providing a mathematical `introduction' and illustration of the `methods' that will be used (in chapter \ref{chap_BD}) and generalized (in chapter \ref{chap_infD_processes}) in a biological context to get `results'.
		\item Chapter \ref{chap_BD} develops a formalism for describing the evolution of finite fluctuating populations of individuals that come in arbitrarily many `types' which vary in a \emph{discrete} character. This yields equations that generalize the classic Price equation and replicator-mutator equation to finite fluctuating populations.
		\item Chapter \ref{chap_infD_processes} extends the ideas developed in chapter \ref{chap_BD} to populations that vary in a single one-dimensional \emph{quantitative} character and derives a so-called `stochastic field theory' that describes evolution in such populations. This also results in some mathematical equations that may be of independent interest to physicists and applied mathematicians.
		\item Chapter \ref{chap_unification} provides a technical summary of the major results by deriving three important stochastic differential equations. These are equations for type frequencies, the population mean value of an arbitrary type-level quantity, and the population variance of an arbitrary type-level quantity, and respectively generalize the replicator-mutator equation, the Price equation, and \cite{lion_theoretical_2018}'s variance equation to finite, stochastically fluctuating populations. These equations predict a directional evolutionary force called `noise-induced selection' that is only seen in finite, fluctuating populations. Some implications for social evolution and community ecology are discussed. I also briefly discuss the field equation formalism I develop in chapter \ref{chap_infD_processes}.
		\item Chapter \ref{sec_disc} provides a quick summary of the major results and discusses biological implications, connections with previous studies, and opportunities for future work. This chapter has no equations (!). Thus, readers who do not want any mathematical equations can skip all other chapters and directly read chapter \ref{sec_disc} if they are interested only in the final results and takeaways. 
\end{itemize}
\section{A more expanded outline of the rest of this thesis}

@@

While the mathematics of chapter \ref{chap_BD} is standard and well-understood, it has, to the best of my knowledge, not been used in this context in the generality we use here. Several specific models of specific systems do use these mathematical techniques, but these papers are often written assuming familiarity with notions in physics and/or mathematics and thus may not be very accessible to theoretical ecologists who do not have formal training in these subjects (but see~\cite{czuppon_understanding_2021} for a recent pedagogical review on the general approach applied to Wright-Fisher and Moran processes, where total population size is constant). As such, chapters \ref{chap_math_background} and \ref{chap_BD} together also serve as a tutorial and technical introduction to some theoretical ideas: For ecologists, the chapter introduces `system size expansions' and illustrates their use in a general setting, and can be seen as a tutorial on modelling finite populations analytically with minimal assumptions; For population geneticists, the chapter illustrates how system-size approximations (`diffusion approximations' in the population genetics literature) can be carried out without assuming a constant (effective) population size and how this generalization has important consequences for the evolutionary forces at play; For physicists and applied mathematicians, the chapter presents a study of the consequences of applying the system-size expansion to the kind of density-dependent birth-death processes that are widely applicable in ecology and evolution - Unlike many physical systems, though demographic processes like birth and death over ecological timescales are usually formulated in terms of population numbers or densities, predictions in evolution are typically in terms of frequencies of types, and this fact has subtle consequences that can be overlooked if one only works with densities.

Chapter \ref{chap_infD_processes} introduces a function-valued process to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via an analog of the system-size approximation to arrive at a `functional' Fokker-Planck equation in which derivatives are replaced by functional derivatives. I show that classic equations from quantitative genetics such as Kimura's infinite alleles model and Lande's gradient dynamics can be derived as the infinite population limit of this stochastic process. I also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation that can be analyzed for specific systems as required. Unlike the systems studied in Chapter \ref{chap_BD}, formalizing the study of the kind of processes we study in Chapter \ref{chap_infD_processes} is an active area of mathematical research~\citep{carmona_stochastic_1999,da_prato_stochastic_2014,prevot_concise_2007,liu_stochastic_2015,bogachev_fokker-planck-kolmogorov_2015,balan_gentle_2018} and the mathematics itself is far from settled.
